The effects of context inference on motor learning: savings, de-adaptation and spontaneous recovery

### 2021-07-07 adaptation/article/article.tex
Added model description

\documentclass[a4paper,doc,floatsintext,natbib]{apa6}
% \documentclass{article, a4paper}
\usepackage[font=small]{caption}
\usepackage{lscape}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage[utf8]{inputenc}
\usepackage{nameref}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{soul}

@@

@@

We can categorize experiments performed with human participants by looking at the amount of contextual information that is available. In some experiments [EGssss], the context is clearly cued to the participant, either in the form of visual cues on in changes to the location of the task. We call these cued-context experiments. In other experiments, partial information is available to participants [EGsss]; we refer to these as partially-cued experiments. While there are no experiments in which no contextual information is available, there are some experiments which contain blocks in which this is the case and participants have no way of inferring the current context. Abusing our notation, we call these experiments (or rather, these sections of the experiments) uncued experiments. In this section, we focus on the two former categories, while the latter is left for the next sections.

@@

@@

\bibliography{../MyLibrary}

>>>>

\documentclass[a4paper,doc,floatsintext,natbib]{apa6}
% \documentclass{article, a4paper}
\usepackage[font=small]{caption}
\usepackage{lscape}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage[utf8]{inputenc}
\usepackage{nameref}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{soul}
\usepackage{todonotes}

@@

\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\argmax}{argmax}

@@

We can categorize experiments performed with human participants by looking at the amount of contextual information that is available. In some experiments [EGssss], the context is clearly \todo{Maybe not use the word 'cue', as many experiments don't have explicit sensory cues}cued to the participant, either in the form of visual cues on in changes to the location of the task. We call these cued-context experiments. In other experiments, partial information is available to participants [EGsss]; we refer to these as partially-cued experiments. While there are no experiments in which no contextual information is available, there are some experiments which contain blocks in which this is the case and participants have no way of inferring the current context. Abusing our notation, we call these experiments (or rather, these sections of the experiments) uncued experiments. In this section, we focus on the two former categories, while the latter is left for the next sections.

@@

\section{Methods}
\subsection{Model description}
We first describe the general principle behind the model, followed, in the next
subsection, by the full mathematical description of each one of its components.

One of the strenghts of this model is that its general principle does not
depend on the specifics of the experiment to which it is applied. However, for
clarity, we will connect the different parts of the model to different
experiments.

This model is a Bayesian alternative to the N-state learners from
\citep[e.g.][]{Lee_Dual_2009}. In this model, there are no explicit different
timescales and multiple states; instead, the model relies on Bayesian learning
for motor adaptation, combined with inference over the current context.

Before proceding with the model description, let us introduce nomenclature:
\begin{itemize}
\item $t$: Time at the moment an observation is made, before inference is done.
\item $\theta_t$: Observation. Note that this is assumed to be a noisy
observation of the generalized state (see below) given by $\theta_t = h(z_t)$.
\item $s_t$: Hidden state, e.g. the position of the hand in reaching
experiments. It does not depend on the context.
\item $a_t$: Action taken after observing $\theta_t$ and inferring the context.
\item $c_t$: Context inferred after observing $\theta_t$.
\item $\omega_t$: Contextual information. This term includes, e.g. visual cues,
proprioceptive feedback, performance feedback.
\item $z_t$: Generalized state, $z_t = \{s_t, c_t, \omega_t\}$. Observed noisily
as $\theta_t$.
\end{itemize}

Context inference can work through context-specific (possibly unreliable) cues,
or through the dynamics of the system (e.g. expected outputs of motor
commands), using prediction error as the force behind inference. In this case,
the generative model can is written as follows:
\[
p(z_t, z_{t-1}, a_{t-1}, \theta_t, \theta_{t-1}) = p(\theta_t | z_t)p(z_t|\theta_{t-1}, a_{t-1})
\]
and inferring the current (hidden) generalized state is done with Bayes' theorem:
\[
  q(z_t | \theta_t, \theta_{t-1}, a_{t-1}) \propto p(\theta_t | z_t)p(z_t|\theta_{t-1}, a_{t-1})
\]

Additionally, after having inferred the current state, motor adaptation is given by
\[
q(\gamma | z_t, z_{t-1}, a_{t-1}) \propto p(z_t | z_{t-1}, a_{t-1}, \gamma)p(z_{t-1})p(\gamma)
\]

where $\gamma$ are the parameters of the internal model for the dynamics,
i.e. the function $f: (z_t, a_t) \rightarrow z_{t+1}$, which determines the beliefs of
the agent regarding how the system evolves after having taken action $a_t$.

Having inferred the generalized state (including context), a decision is made
sampling from a distribution:
\begin{equation}
p(a_t | z_t, \beta) = \displaystyle\sum_{m \in M}q(C_t)p(a_t | z_t, \beta, m)
\end{equation}
where $M$ is the set of all relevant forward models (see below), and
$p(a_t | ... m)$ is the probability distribution over actions prescribed by the
forward model $m$.

In the following sections, we present a specific mathematical implementation
of the principles outlined above. This is the implementation used throughout the
simulations in the main text.

\subsubsection{Mathematical implementation}
The generalized state is given by:
\begin{equation}
z_t = \{x, \theta_t, \gamma_t, \omega_t\}
\end{equation}
where $x$ is, e.g. in a reaching task, the current position of the hand in
Cartesian coordinates, where the origin is the starting point. We assume that
motor commands are issued every $\Delta t$ for simplicity. At the beginning of
each time interval, the context is inferred combining the contextual
information and the prediction error of the outcome of the previous motor
command. This yields:
\begin{align}
  q(C_t) &= q(C_t | \omega_t)q(C_t | s_t, s_{t-1}, a_{t-1}) \\ \label{eqn:estimated-context}
  q(C_t | \omega_t) &\propto p(\omega_t | C_t)p(C_t) \\
  q(C_t | x_t, x_{t-1}, a_{t-1}) &\propto p(x_t | C_t, x_{t-1}, a_{t-1})p(x_t)p(x_{t-1})p(C_t)
\end{align}
The terms $p(C_t)$ refer to the prior probability of the context at the
beginning of the trial: at the first trial, this refers to prior beliefs over
which contexts are more common/likely, which we call $p_0(C_t)$. At each
subsequent trial, it incorporates the belief that has so far been accumulated,
given previous trials. Note that certain events, such as the start of a new
block of trials (see main text) can return an agent to a state in which
$p_0(C_t)$ becomes relevant.

We do not directly model hand movements or eye position. Instead, we model
systems in the motor error space: the position $s_t = 0$ represents a trial in
which no motor error occurred. We make use of the fact that in most
experiments, motor adaptation needs to happen in one of two directions
(e.g. clockwise vs. counter-clockwise) to keep the error space
two-dimensional. With this, $s_t > 0$ represents motor error in one direction
(e.g. clockwise) and $s_t < 0$, error in the other direction. We made this
choice because, in this work, we would model different experiments with
different state spaces; but all these experiments have the commonality of
two-dimensional error spaces. In order to model a specific experiment in its
own physical space (e.g. hand positions in a reaching task), only the dynamical
system of \eref{eqn:dynamical-system} would need to be adapted.

We assume that the decision-making agent has an internal forward model that
maps motor commands to outcomes:
\begin{equation}
p(s_{t+1} | s_t, a_t, m) = f_m(s_t, a_t)
\end{equation}
where $a_t$ are motor commands (actions). We further assume that participants
have learned $f_m$ throughout their lives and use it to decide which motor
command to issue. When a perturbation is introduced, to which the participant
must adapt, we assume they learn a new forward model $f_m^{\phi}$, where $\phi$
represents the parameters of the function that participants must learn. For
example, in a saccade task, $\phi$ represents the new gain. Note that $\phi$
depends on the context and participants need to use and update multiple
forward adapted models $f_m^{\phi}$ throughout one experiment.

Given a forward model $f_m$ (or $f_m^\phi$), action selection is done by
building a probability distribution over possible actions $a_t$ based on how
likely they are to produce zero motor error. For example, in saccade
experiments, motor commands are issued based on how likely they are to move the
fovea to the target. These motor commands are issued to counteract intrinsic
sources of error, such as muscle variability \todo{expand this?}, and extrinsic
sources of error, such as the experimental manipulations on saccade gain. Thus,
the expected observed error can be written as:
\begin{equation}
p(\epsilon | a_t) \propto p(\epsilon_{\text{intrinsic}} | a_t) + p(\epsilon_{\text{extrinsic}} | a_t)
\end{equation}
and a motor command can be chosen to maximize the chance of observing zero error:
\begin{equation}
p(a_t) = argmin_{a_t} p(\epsilon | a_t)
\end{equation}

For simplicity, we assume that motor errors are Gaussian, such that:
\begin{equation}
p(\epsilon | a_t) = N(\mu_{\text{extrinsic}}, \sigma_\epsilon)
\end{equation}
where $\sigma_\epsilon$ incorporates both intrinsic and extrinsic sources of
error. The mean of the error is assumed to have no component, as motor
commands, in the absence of external errors, should produce an error centered
at zero [CITATION MISSING].

With this, a forward model will produce a motor command with outcomes centered
at $-\mu_{\text{extrinsic}}$, which counteracts the externally-induced errors.

To complete the action-selection picture, we return to context inference. As
discussed above, action selection is made via a weighted sum of the different
forward models:
\begin{equation}
p(a_t | ...) \propto \displaystyle \sum_{m \in M} q(C_i)p(a_t | m)
\end{equation}
From this distribution, an action is sampled at each trial.

To update the agent's estimates on the parameters of the forward models, we
make use of Bayes' theorem with conjugate priors. This approach has the
advantage of not only greatly simplifying computational calculations, but also
presenting update equations with great intuitive appeal.

In the following equations, we drop the dependence on the context and
observations for notational simplicity:
\begin{equation}
q(\phi | s_{1:t}) \propto p(s_{1:t} | \phi)p(\phi)
\end{equation}
where $s_{1:t}$ are the observed states up until time $t$. The likelihood will
be a Gaussian:
\begin{align}
  p(s_t | \theta) &= N(\mu_x, \sigma_x)  \\
  \mu_x &= \theta_{t-1} + x \\
  \sigma_x &= \xi_0 1
\end{align}
This likelihood is the probability of observing $x$ given the previous estimate
of the parameters of the force field $\phi_{t-1}$, and given the dynamics of
the system (see below).

To update $\phi$, we assumed that for each context, the agent's belief over
the magnitude of the adaptation is given by a normal distribution:
\begin{equation}
p(\phi) = N(\mu_\phi, \sigma_\phi) \label{eqn:data-dist}
\end{equation}
where $\mu_\phi$ and $\sigma_\phi$ are parameters to be estimated at each
trial. Before the experiment begins, the agent will have a prior distribution
over these parameters. A standard Bayesian approach is to choose NormalGamma
priors:
\begin{equation}
p(\mu_\phi, \sigma_\phi) = NG(\mu_0, \nu_0, \alpha_0, \beta_0)
\end{equation}
where $\mu_0, \nu_0, \alpha_0$ and $\beta_0$ are free hyperparameters of the
model. We assume the following hyperparameter values \todo{Justify}:
\begin{align} \mu_f^{(0)} &= (0, -1, 1) \\
  \nu_f^{(0)} &= 0.1 \\
  \alpha_f^{(0)} &= 1 \\
  \beta_f^{(0)} &= 0.5
\end{align}
Note that $\mu_f^{(0)}$ is different for each context and is thus provided as a
vector $(0, -1, 1)$ representing the baseline and two opposing adaptations
(e.g. clockwise and counter-clockwise visuomotor rotations), respectively. For
all other hyperparameters, all contexts have the same value. These values
ensure that the prior over the hyperparameters $(\mu_\phi, \sigma_\phi)$ has the mode
at $(\mu_\phi^{(0)}, 1)$.

The update equations for the magnitud parameters are given by:
\begin{equation}
q(\mu_f, \sigma_f | x_t, x_{t-1}, a_{t-1}) \propto p(x_t |
x_{t-1}, a_{t-1}, C_t)p(\mu_f, \sigma_f) \label{eqn:context-from-x}
\end{equation}
where $p(x_t | x_{t-1}, a_{t-1}, C_t)$ is a Gaussian distribution centered
around $\mu_a$ with standard deviation $\sigma_a$, which is a free parameter of
the model. It is assumed that the standard deviation over outcomes ($\sigma_a$)
is related to that of the parameter $\mu_f$ as follows:
\begin{equation}
\sigma_f = \sigma_a / \nu_f
\end{equation}
which makes $\nu_f$ a parameter to estimate. Each one of the forward models can
be updated at every trial, depending on context inference (see main text). The
updated parameters for any one forward model $i$ are given by:
\begin{align}
  \mu_\phi^{(t)} &= \frac{\nu_\phi^{(t-1)} \mu_\phi^{(t-1)} + q(C_i)s_t}{\nu_\phi^{(t-1)} + q(C_i)} \\
  \nu_\phi^{(t)} &= \nu_\phi^{(t-1)} + q(C_i) \\
  \alpha_\phi^{(t)} &= \alpha_\phi^{(t-1)} + q(C_i) / 2 \\
  \beta_\phi^{(t)} &= \beta_\phi^{(t-1)} + \frac{q(C_i)\nu_\phi^{(t-1)}}{\nu_\phi^{(t-1)} +
                  q(C_i)}\frac{\left(s_t - \mu_\phi^{(t-1)}\right)^2}{2}
\end{align}
where $q(C_i)$ is the posterior probability of context $C_i$ from
\eref{eqn:estimated-context}. Note that when $q(C_i) = 0$, the model $i$ is not
updated.

The effect of the hyperparameters of the priors are worth a note or two, as
they are complex.

Naturally, $\mu$ affects the initial estimate of the adaptation, in the same
units as the necessary adaptation. $\nu$ encodes how stable this hyperprior is:
higher values (e.g. 10000) all but guarantee that the hypermean will not move
in the face of evidence; In principle, enough evidence should still move it,
but that won't happen during an experiment. Smaller values (e.g. 1 / force\_sd,
as is the default on the code) make the hypermean (and thus the mean) follow
evidence more freely. Note that as more observations are accumulated, $\nu$
becames bigger and bigger, solidifying the value of the hypermean.

The hyperparameters $\alpha$ and $\beta$ are a bit more complex. Note that the
mean of a Gamma distribution is $\beta / (\alpha \nu)$; this mean is being used
as the standard deviation of a Gaussian by the rest of the agent (cheating),
which makes it an important measure of uncertainty. While setting the default
hyperparameters, the values used are $\alpha = 0.5 / \sigma_0$ and
$\beta = 0.5$, where $\sigma_0$ is the \textit{a priori} estimate of the
standard deviation of the force exerted by the environment (force\_sds in the
code), which controls the initial learning rate. This makes the initial
standard deviation equal $\sigma_0$, which makes it consistent with the
fixed-force model in its interface. The 0.5 values ensure that uncertainty is
large at the beginning and during the experiment is greatly reduced, but never
to a point where it's ``visually'' too small. Changing this 0.5 would make the
standard deviation change more quickly, making the model more or less precise
in its predictions, independently of the volatility of the mean of the
adaptation (via the hypermean).

The baseline model defaults to different values that make it a lot more
stable. The hyperstd of the mean is set to 10,000, which makes the mean
entirely stable during the duration of the experiment. The values of $\alpha$
and $\beta$ are fixed regardless of $\sigma_0$ such that the standard deviation
is 0.001 (compare that to the size of the adaptations, around 0.0125), and the
hyperparameters of the standar deviation are stable during the experiment.

@@

\bibliography{../../MyLibrary}
